Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-05-11 manuscript/manuscript.Rmd
added some more references.
minor edits of the first section.

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. The controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007] Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were invovled. Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023].

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology.

>>>>

The now extensive corpus of experimental archaeology have principally focusesd on the generation of knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

1) True controlled experiments are rare. Randomized controlled experiments *sensu stricto* are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from datasets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023]. It has been a debatable issue in philiosophy of science whether RCT represents the golden standard of knowledge [@cartwright2007].
2) Trade-off between causality and generalizability in experimental design. For example, in a series of recently published experimental studies aiming at understanding the role of language in the transmission of lithic technologies, standardized procedures like video teaching or using bricks as raw materials are very often.

In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. This also echoes the post-positivist turn in psychology, a field that is dominated by experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].
